Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2023-08-28 noise_covid_v2.Rmd
edit thetadiff surface
edit text; move code for creating surface

---
title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
author: "Rachael Aber"
header-includes: \usepackage{setspace}\doublespacing
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
fontsize: 12pt
---

@@

Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.

Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable <-(exclude this sentence?**). We present a method to identify shifts in population-level dispersion in incidence, so the impact of events on superspreading dynamics can be assessed using freely available population-level data.

@@

Naively, the negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992). However, issues arise when modeling the entire time series as negative binomial - namely, that autocorrelation remains unaccounted for (Barron 1992).

Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions. Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al. 2002). The final process potentially underlying overdispersed incidence, varying population effective reproduction number, has been studied extensively in investigations into seasonal forcing of population-level effective reproduction number.

@@

Approximately one month of data on either side of November 26th, 2020 was used in the analysis. This step was done to include a set "wave" of incidence in the model fit to each side of the hypothesized change. However, we acknowledge that there is a trade-off between including fewer waves and increasing the power of the proposed test by including more time series data. In both the full data and the subset of data, rows that were all zeroes were omitted.

@@

A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009). The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time to account for autocorrelation in case counts. Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019). The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 30 days. In addition, increasing degrees of freedom above three results in overfitting to the time series and thus potentially insignificant changes in dispersion are more readily detected.

@@

In this case, we use a log link function:
$$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
  $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
  $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$

@@

To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion is present under this model and found that only a few dispersion statistics were less than or equal to one. Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1.

```{r}
load("data/dispersion_stats.Rdata")
length(which(disp_stats > 1))/length(disp_stats)
```

Directly comparing between the quasi-Poisson and the Poisson models via deviance residuals is not possible because the deviance residuals are identical. Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible (Ver Hoef and Boveng 2007). In the quasi-Poisson model (called the linear negative binomial specification by Barron (1992)), the variance is a linear function of the mean, with a slope of one representing the Poisson special case. By contrast, the negative binomial model may more appropriately represent the data when the variance has approximately a quadratic relationship with the mean. Since the true mean and variance of the data set is unknown, one could approximate these quantities with estimates $\hat{\mu}$ by and $\widehat{\sigma^2}$, respectively. These quantities should be estimated using both the Poisson and the negative binomial model. Variance can be estimated by averaging squared residuals in each fitted mean category. The number of fitted mean categories (bins) should be chosen such that there are an adequate number of observations informing each variance estimate, and also an adequate number of variance estimates to visualize whether a trend is present. Visualization of individual squared residuals is also possible, but these may not be reliable estimates of the variance.

We use a diagnostic plot approach in which we rely on the expected value of the residuals to infer "p" using the slope of a regression of log(Y) on log(X), where X is r^2 - Y. The expectation of X is phi * mu^p, so
 log(E(X)) ~ log(phi) + p log(E(Y))

![Diagnostic Plot](figures/diagnostic_plot.pdf)

@@

![Empirical change in dispersion](./figures/empirical_increase.pdf)

@@

For every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$. A test of the two-sided alternative that $\theta$ before is different from theta after (different theta after Thanksgiving) was performed.

@@

$$ \int_0^\infty {(n/10)/\sqrt{2\pi(\sigma^2))} e^{-(t-61)^2/(2\sigma^2)}}dt = n/10$$
So, we have the desired Gaussian function:
$$ \mu_t = (n/10)/\sqrt{2\pi\sigma^2} e^{-(t-61)^2/(2\sigma^2)} $$
So that integration will result in a final outbreak size of n/10.

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed. 
The resulting probabilities of rejection are visualized below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true. We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (see Supplement).

@@

![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf)

@@

@@

![LRT  p-values v. Theta Difference](figures/pvals_sim_LRT.pdf)

@@

![Mapping the Thanksgiving hypothesis across the US](./figures/map_difftheta_US.pdf)

![LRT Scan Figure](./figures/LRT_scan_figure.pdf)

![The LRT surface](./figures/lrt_surface_figure.pdf)

![The Theta Difference Surface](./figures/thetadiff_surface.pdf)

In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.

>>>>

---
title: Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model
author: "Rachael Aber, Yanming Di, Ben Dalziel"
header-includes: \usepackage{setspace}\doublespacing
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
fontsize: 12pt
---

@@

Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of events/gatherings (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes in the US or even mean changes per county.

Implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable. We present a method to identify shifts in population-level dispersion in incidence, so the impact of events or NPIs on superspreading dynamics can be assessed using freely available population-level data.

@@

Naively, the negative binomial distribution might accurately model a time series if there is a changing process mean: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial (Cook 2009). Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion that all lead to overdispersion (Barron 1992). Improved understanding of the causes of overdispersion in incidence may provide insight into interactions between host population structure and contagion processes, contributing to general predictive understanding in a range of systems, and understanding the limits of predictability.Previously, efforts to quantify the processes underlying overdispersed incidence have included the study of underlying offspring distributions. Also, comparison of simulated and observed incidence time series has been used to estimate the role of environmental and demographic stochasticity in measles: across community sizes, demographic stochasticity of measles becomes more important in small human populations, where dynamics can’t be described as well simply by contact rate and birth rate (Grenfell et al. 2002). The final process potentially underlying overdispersed incidence, varying population effective reproduction number, has been studied extensively in investigations into seasonal forcing of population-level effective reproduction number.

@@

Approximately one month of data on either side of November 26th, 2020 was used in the analysis. This step was done to include a set "wave" of incidence in the model fit to each side of the hypothesized change. However, we acknowledge that there is a trade-off between including fewer waves and increasing the power of the proposed test by including more time series data.

We additionally use the full dataset to "scan" for variability breakpoints using a likelihood ratio test statistic.

In both the full data and the subset of data, rows that were all zeroes were omitted.

@@

Issues arise when modeling the entire time series as negative binomial - namely, that autocorrelation remains unaccounted for (Barron 1992). A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009). The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time to account for autocorrelation in case counts. Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019). The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 30 days. In addition, increasing degrees of freedom above three results in overfitting to the time series and thus potentially insignificant changes in dispersion are more readily detected.

![Fitted Curves With Three Spline Degrees of Freedom](figures/FL_signif_fittedcurves.pdf){width=30%, height=30%}

@@

In this case, we use a log link function:

$$log(E[Y_i]/n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$
  $$log(E[Y_i])-log(n_i) = \beta_1h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i)$$

  $$log(E[Y_i]) = \beta_1(h_1(t_i) + \beta_2h_2(t_i) + \beta_3h_3(t_i) + log(n_i)$$

@@

```{r}
load("data/dispersion_stats.Rdata")
```

To identify an appropriate conditional distribution, the value of Poisson model residual deviance (D) divided by its degrees of freedom (df) was used to verify that overdispersion is present under this model and found that only a few dispersion statistics were less than or equal to one. Since residual deviance of a model is expected to be asymptotically distributed as $\chi^2_{df}$, where the degrees of freedom are the number of observations minus the number of estimated parameters, an indication that a model may be underestimating dispersion (that the data is inconsistent with the model) is that D/df >> 1. The proportion of dispersion statistics that are greater than one in US counties over the Thanksgiving period is `r length(which(disp_stats > 1))/length(disp_stats)`.

Directly comparing between quasi-Poisson and Poisson models via deviance residuals isn't possible because the deviance residuals are identical. Therefore, to evaluate whether the negative binomial conditional distribution is needed (opposed to a Poisson/quasi-Poisson conditional distribution with the same model of process mean), inspection of mean-variance relationships using a diagnostic plot is possible (Ver Hoef and Boveng 2007). In the quasi-Poisson model (called the linear negative binomial specification by Barron (1992)), the variance is a linear function of the mean, with a slope of one representing the Poisson special case. By contrast, the negative binomial model may more appropriately represent the data when the variance has approximately a quadratic relationship with the mean. Since the true mean and variance of the data set is unknown, one could approximate these quantities with estimates $\hat{\mu}$ by and $\widehat{\sigma^2}$, respectively. These quantities should be estimated using both the Poisson and the negative binomial model. Variance can be estimated by averaging squared residuals in each fitted mean category. The number of fitted mean categories (bins) should be chosen such that there are an adequate number of observations informing each variance estimate, and also an adequate number of variance estimates to visualize whether a trend is present. Visualization of individual squared residuals is also possible, but these may not be reliable estimates of the variance.

We use a diagnostic plot approach in which we rely on expected value of the residuals to infer "p" using the slope of a regression of log(Y) on log(X), where X is residual^2 - Y.

$$E[X] = \phi * \mu^p$$
$$log(E(X)) \sim log(\phi) + p*log(E(Y))$$

![Diagnostic Plot](figures/diagnostic_plot.pdf){width=30%, height=30%}

We use a second diagnostic plot, squared residuals versus squared fitted values to look for a linear trend in the Florida counties with a significant LRT test result for the Thanksgiving hypothesis. The constant dispersion model (NB2) model is appropriate if we observe an approximately linear trend.

![Diagnostic Plot 2](figures/diagnostic_plot.2.pdf){width=30%, height=30%}

@@

![Empirical change in dispersion](./figures/empirical_increase.pdf){width=30%, height=30%}

@@

For every county, the model was fit separately to either side of November 26th, 2020 in order to compute values of $\hat{\theta_1}$ and $\hat{\theta_2}$. A test of the two-sided alternative that $\theta$ before is different from $\theta$ after (different theta after Thanksgiving) was performed.

@@

$$ \int_0^\infty {(n/10)/\sqrt{2\pi(\sigma^2))} e^{-(t-61)^2/(2\sigma^2)}}dt = n/10$$
So, we have the desired Gaussian function:

$$ \mu_t = (n/10)/\sqrt{2\pi\sigma^2} e^{-(t-61)^2/(2\sigma^2)} $$
So that integration will result in a final outbreak size of n/10.

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed. 
The resulting probabilities of rejection are tabulated below: the proportion of tests that reject is around 0.05 for those instances where the null is true (the probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true. We also simulated data assuming that theta was a function of the mean of the process in order to assess the robustness of the test (Supplement).

@@

Additionally, the relationship between the p-values resulting from the Wald test and the effect size is as expected.

![Wald Test p-values v. Theta Difference](figures/pvals_sim_Wald.pdf){width=30%, height=30%}

@@

The tabulated Type I error rate and power, as well as the relationship between p-values and effect size is presented below for the likelihood ratio test approach.

@@

![LRT  p-values v. Theta Difference](figures/pvals_sim_LRT.pdf){width=30%, height=30%}

@@

The spatial pattern of theta difference after Thanksgiving 2020 shows that there is spatial correlation in the degree of change. In particular, spatial clusters of increased dispersion occur on the East Coast and in the South.

![Mapping the Thanksgiving hypothesis across the US](./figures/map_difftheta_US.pdf){width=30%, height=30%}

Additionally, the likelihood ratio test statistic does not necessarily track with changes in mean incidence in a county, but rather captures dispersion changes. In some time windows, the Likelihood-Ratio statistic can't be computed due to zero variance: the variance of mu is a function of mu and dispersion, both of which are zero in a stretch of zeroes.

![LRT Scan Figure](./figures/LRT_scan_figure.pdf){width=30%, height=30%}

In visualizing a surface of the likelihood ratio test statistic and associated direction of the change in theta, it appears that

![The LRT Surface](./figures/lrt_surface_figure.pdf){width=30%, height=30%}

![The Theta Difference Surface](./figures/thetadiff_surface.pdf){width=30%, height=30%}
